Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2017-06-21 paper/submitted/round2_png/algorithms_v4_cmame.tex
more edits

	$\epsilon_0=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF_0})(\Lambda_0)+\mathcal M_{HF}(\Psi_{MF_0})-\mathcal M_{MF_0}(\Psi_{MF_0})$}
	\State{Calculate QoI $I(q_{MF_0},u_{MF_0})$}
	\While{$i<$ \texttt{maxIter} and $|\epsilon_i/(\epsilon_i+I(q_{MF_i},u_{MF_i}))|>$ \texttt{errTol}}
		\State{\begin{varwidth}[t]{\linewidth}Localize $\epsilon_i$ (see \Cref{sec:errLocal}) and use this decomposition to guide \par\hskip\algorithmicindent formation of new mixed-fidelity model MF$_{i+1}$\end{varwidth}}
		\State{$i\gets i+1$}
		\State{Solve for stationary point $\Psi_{MF_i}$ of augmented Lagrangian $\mathcal{M}_{MF_i}$}
		\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_i}$, for

@@

		$\quad\quad \epsilon_i=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF_i})(\Lambda_i)+\mathcal M_{HF}(\Psi_{MF_i})-\mathcal M_{MF_i}(\Psi_{MF_i})$}
		\State{Calculate QoI $I(q_{MF_i},u_{MF_i})$}
	\EndWhile \\
\Return{model MF$_i$ and QoI estimate $I(q_{MF_i},u_{MF_i})$}
\EndProcedure
\Statex
\end{algorithmic}
\end{algorithm}
%

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh refinement, though in this case the method described in~\cite{BecVex05} would likely be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.

@@

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. Recall the error estimate 
%
\begin{equation}
\epsilon=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF})(\Lambda)+\mathcal M_{HF}(\Psi_{MF})-\mathcal M_{MF}(\Psi_{MF})
\end{equation}
%
from \Cref{alg:refSeries}. Let $(Q^h\times U^h\times U^h)^3 \subset (Q\times U\times U)^3$ be the finite-dimensional conforming subspace in which we solve for the approximations $\Lambda^h$ and $\Psi_{MF}^h$; we seek to decompose the error estimate
%
\begin{equation}
\epsilon^h=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF}^h)(\Lambda^h)+\mathcal M_{HF}(\Psi_{MF}^h)-\mathcal M_{MF}(\Psi_{MF}^h).
\end{equation}
%
Define a basis $\Phi^h=\{(\varphi,\upvarphi)_i\}_{i\in I}$ consisting of locally supported functions such that $\textrm{span}(\Phi^h)=(Q^h\times U^h\times U^h)^3$, so that we can write $(\Lambda^h,\Psi_{MF}^h)=(\sum\limits_{i\in I}\varphi_i\lambda_i,\sum\limits_{i\in I}\upvarphi_i \psi_i)$. Then the error estimate satisfies
%
\begin{equation}
\epsilon^h \leq \sum_{i\in I} \varepsilon^h_i,
\end{equation}
%
where,
%
\begin{equation}\label{eq:basisblame}
\varepsilon^h_i = \left| -\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF})(\lambda_i\varphi_i)+\mathcal M_{HF}(\upvarphi_i \psi_i)-\mathcal M_{MF}(\upvarphi_i \psi_i) \right|
\end{equation}
%
can be interpreted as the error contribution from the basis function $(\varphi,\upvarphi)_i$. Near the interfaces between the low-fidelity and high-fidelity regions, basis functions may have their support divided between the two regions and thus have a nonzero error contribution. The basis functions with the largest error contributions are flagged, and the elements in their support are added to the high-fidelity subdomain. Basis functions whose support lie completely in the high-fidelity region will have a zero contribution and the elements in their support are never flagged for refinement again.

>>>>

	$\epsilon_0=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF_0})(\Lambda_0)+\mathcal M_{HF}(\Psi_{MF_0})-\mathcal M_{MF_0}(\Psi_{MF_0})$}
	\State{Calculate QoI $I(q_{MF_0},u_{MF_0})$}
	\While{$i<$ \texttt{maxIter} and $|\epsilon_i/(\epsilon_i+I(q_{MF_i},u_{MF_i}))|>$ \texttt{errTol}}
		\State{Localize $\epsilon_i$ (see \Cref{sec:errLocal}) and use this decomposition to guide \par\hskip\algorithmicindent formation of new mixed-fidelity model MF$_{i+1}$}
		\State{$i\gets i+1$}
		\State{Solve for stationary point $\Psi_{MF_i}$ of augmented Lagrangian $\mathcal{M}_{MF_i}$}
		\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_i}$, for

@@

		$\quad\quad \epsilon_i=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF_i})(\Lambda_i)+\mathcal M_{HF}(\Psi_{MF_i})-\mathcal M_{MF_i}(\Psi_{MF_i})$\label{alg-line:MFerrorest}}
		\State{Calculate QoI $I(q_{MF_i},u_{MF_i})$}
	\EndWhile \\
\Return{model MF$_i$ and QoI estimate $I(q_{MF_i},u_{MF_i})$}
\EndProcedure
\Statex
\end{algorithmic}
\end{algorithm}
%

As an example, consider a pair of low- and high-fidelity models described by the convection-diffusion and convection-diffusion-reaction equations
\begin{subequations}
\begin{align}
\textrm{low-fidelity:   } & k_d\nabla^2 u - \vec{V}\cdot\nabla u = q, \\
\textrm{high-fidelity:   } & k_d\nabla^2 u - \vec{V}\cdot\nabla u + k_ru^2 = q,
\end{align}
\end{subequations}
where $k_d$ and $k_r$ are diffusion and reaction coefficients, respectively, and the state $u$ and parameters $q$ are continuous functions over a domain $\Omega$ with homogenous Dirichlet boundary conditions. To form the mixed-fidelity models, we divide the domain into complementary subdomains, $\Omega_{HF}$ and $\Omega_{LF}$, where the high- and low-fidelity models are solved, respectively. The resulting mixed-fidelity models can be described by,
%
\begin{equation}
k_d\nabla^2 u - \vec{V}\cdot\nabla u + k^{MF}_ru^2= q,
\end{equation}
%
where $k^{MF}_r$ is a piecewise-constant reaction coefficient that's zero in the $\Omega_{LF}$ and equal to $k_r$ in $\Omega_{HF}$; initially, $\Omega_{LF}=\Omega$. Let the QoI be described by $I(q,u)=\int_{\Omega_I\subset\Omega} u \:\textrm{d}\Omega$, where $\Omega_I$ is the QoI region. Let the observations be of the state at some set of locations, so that the data-misfit term is $\frac{1}{2}\int_\Omega \delta_{obs}(u-d)^2\:\textrm{d}\Omega$, where $\delta_{obs}$ is a sum of Dirac delta functions indicating the locations of the observations. Let the regularization term be $R(q)=\frac{\beta}{2}\int_\Omega \|\nabla q\|_2^2\:\textrm{d}\Omega$, where $\beta$ is a regularization coefficient. Assuming divergence-free velocity $\vec{V}$, the mixed-fidelity model has augmented Lagrangian
\begin{eqnarray}
\mathcal{M}_{MF}((q,u,z)(p,v,y)) &=& \int_{\Omega_I\subset\Omega} u \:\textrm{d}\Omega \nonumber\\
&+& \int_\Omega p(-\beta\nabla^2q+z) \nonumber\\
&+& v(k_d\nabla^2z+\vec{V}\cdot\nabla z+2k^{MF}_rzu+\delta_{obs}(u-d)) \nonumber\\ 
&+& y(k_d\nabla^2u-\vec{V}\cdot\nabla u+k^{MF}_ru^2+q)\:\textrm{d}\Omega.
\end{eqnarray}
Taking variations with respect to the auxiliary variables $(p,v,y)$ recovers the equations for the primary variables; taking variations with respect to the primary variables gives the equations for the auxiliary variables. Solving for the primary variables and then the auxiliary variables gives the stationary point $\Psi_{MF}$ of $\mathcal{M}_{MF}$. The weighted residual of the high-fidelity primary and auxiliary equations, evaluated at $\Psi_{MF}$, describes a linear functional in the weights (see \Cref{eq:supadjout}). Taking the adjoint of these high-fidelity primary and auxiliary systems, with output described by this linear functional, gives the supplementary adjoint; the components of the supplementary adjoint corresponding to the primary and auxiliary systems can be solved separately. Having obtained $\Psi_{MF}$ and the supplementary adjoint $\Lambda$, we evaluate the error estimate (see Step \ref{alg-line:MFerrorest} of \Cref{alg:refSeries}), localize it (see \Cref{sec:errLocal}), and use the localization to guide the selection of elements currently in $\Omega_{LF}$ to include in $\Omega_{HF}$ in the next iteration. We return to this example and give results of applying \Cref{alg:refSeries} to this pair of models in \Cref{sec:cdvcdr}.

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter and/or state space. The derived error estimate is not applicable to all models, however. It is assumed that the two models share the same domain; for example, the low-fidelity model can not be a one-dimensional average of a two-dimensional high-fidelity model. \red{The high-fidelity model must by expressible in a weak form}, so this cannot be applied to, for example, a high-fidelity model of chemical reactions using kinetic Monte Carlo. %if the LF model does not have a weak form, say it's from a response surface...what would the components of Psi be?
We need some degree of compatibility between the two models, in that we require the low-fidelity solutions to be interpretable in the context of the high-fidelity model; specifically, we require that $\Psi_{LF}$ be in a space admissible to $\mathcal{M}'_{HF}$ and $\mathcal{M}_{HF}$. For example, compared to the high-fidelity model, low-fidelity model may have a coarser mesh, basis functions of a lower order, or basis functions spanning a reduced subspace. The high-fidelity model could describe changes in states that have a fixed nominal value in the low-fidelity model; the high-fidelity model could allow anisotropy of the inferred parameters while the low-fidelity model could constrain their components and require isotropy. We note that in the case where the two models correspond to two levels of mesh refinement, method described in~\cite{BecVex05} would likely be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead.
%, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$. %implied by admissibility of LF to HF

@@

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. Recall the error estimate 
%
\begin{equation}
\epsilon=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF})(\Lambda)+\mathcal M_{HF}(\Psi_{MF})-\mathcal M_{MF}(\Psi_{MF})
\end{equation}
%
from \Cref{alg:refSeries}. Let $(Q^h\times U^h\times U^h)^3 \subset (Q\times U\times U)^3$ be the finite-dimensional conforming subspace in which we solve for the approximations $\Lambda^h$ and $\Psi_{MF}^h$; we seek to decompose the error estimate
%
\begin{equation}
\epsilon^h=-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF}^h)(\Lambda^h)+\mathcal M_{HF}(\Psi_{MF}^h)-\mathcal M_{MF}(\Psi_{MF}^h).
\end{equation}
%
Define a basis $\Phi^h=\{(\varphi,\upvarphi)_i\}_{i\in I}$ consisting of locally supported functions such that $\textrm{span}(\Phi^h)=(Q^h\times U^h\times U^h)^3$, so that we can write $(\Lambda^h,\Psi_{MF}^h)=(\sum\limits_{i\in I}\varphi_i\lambda_i,\sum\limits_{i\in I}\upvarphi_i \psi_i)$. Then the error estimate satisfies
%
\begin{equation}
\epsilon^h \leq \sum_{i\in I} \varepsilon^h_i,
\end{equation}
%
where,
%
\begin{equation}\label{eq:basisblame}
\varepsilon^h_i = \left| -\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{MF}^h)(\lambda_i\varphi_i)+\mathcal M_{HF}(\upvarphi_i \psi_i)-\mathcal M_{MF}(\upvarphi_i \psi_i) \right|
\end{equation}
%
can be interpreted as the error contribution from the basis function $(\varphi,\upvarphi)_i$. Near the interfaces between the low-fidelity and high-fidelity regions, basis functions may have their support divided between the two regions and thus have a nonzero error contribution. The basis functions with the largest error contributions are flagged, and the elements in their support are added to the high-fidelity subdomain. Basis functions whose support lie completely in the high-fidelity region will have a zero contribution and the elements in their support are never flagged for refinement again.

### 2017-06-21 paper/submitted/round2_png/mathematical_formulation_v4_cmame_vikram_HLedit.tex
more edits

A given physical system can be described to varying degrees of fidelity using different models. One has a choice in how to define the stucture of the inferred parameters $q$, as well as how to define the state equation and variables. For example, the parameter $q$ can be modeled as a single scalar value or as an infinite-dimensional distributed field variable. The state equation can be nonlinear and \red{multiscale}, requiring an expensive computation to resolve (e.g.\ the Navier Stokes equations) or a linear, \red{single physics} model (e.g.\ Laplace's equation) \red{are NS and Laplace's possibly different modeling choices for the same situation?}. All these different modeling choices lead to different versions of the optimization problem described by \Cref{eq:invOpt} that would need to be solved to infer the unknown parameters.

We assume that there exists an `ideal' model for solving the inverse problem, which we will call the high-fidelity model; given adequate computational resources, we would use this high-fidelity to formulate and solve our inverse problem. We use the subscript $_{HF}$ to denote the high-fidelity inverse problem,
%
\begin{subequations}
\label{eq:invOptHF}
\begin{align}
q_{HF},u_{HF} &=&\arg\min\limits_{q,u} & \quad J(q,u)=\frac{1}{2}\|d-C(u)\|_2^2 + R(q), \label{eq:invOpt_objHF} \\
&&\textrm{s.t. }& \quad a_{HF}(u,q)(\phi)=\ell_{HF}(q)(\phi),\quad\forall\phi\in U_{HF} \label{eq:invOpt_consHF}
\end{align}
\end{subequations}
%
where $q_{HF} \in Q_{HF}$ and $u_{HF}\in U_{HF}$ are the inferred parameters and corresponding state. In practice, one has limited computational resources, so one may need to instead infer the parameters using a less computationally demanding model, which we call the low-fidelity model. We use the subscript $_{LF}$ to denote the low-fidelity inverse problem,
%
\begin{subequations}
\label{eq:invOptLF}
\begin{align}
q_{LF},u_{LF} &=& \arg\min\limits_{q,u} & \quad J(q,u)=\frac{1}{2}\|d-C(u)\|_2^2 + R(q), \label{eq:invOpt_objLF} \\
&& \textrm{s.t. }& \quad a_{LF}(u,q)(\phi)=\ell_{LF}(q)(\phi),\quad\forall\phi\in U_{LF}, \label{eq:invOpt_consLF}
\end{align}
\end{subequations}
%
where $q_{LF} \in Q_{LF}$ and $u_{LF}\in U_{LF}$ are the inferred parameters and corresponding state. If there is \red{compatibility (need to elaborate on "compatibility" somehow; was this discussed in the stokes-NS paper? compatibility comes up in fusing the models, and having the same QoI functional apply to variables of all fidelities)} between the low-fidelity and high-fidelity solution spaces for the parameter and state variables, one may combine the two models to form a mixed-fidelity model and corresponding inverse problem. The low-fidelity model can be used in part of the computational domain, and the high fidelity in the remainder; we denote the mixed-fidelity model and its variables with the subscript $_{MF}$. How one chooses to combine the low- and high-fidelity models can be informed by what aspect of the inferred parameters is of interest.

In a {\em goal-oriented inverse problem}, the ultimate purpose of inferring the unknown parameters is to calculate some quantity of interest (QoI). We denote a scalar QoI by a functional that maps the parameters and state to our QoI, $I:Q \times U \to \R$. %The QoIs evaluated with the high ($I(q_{HF},u_{HF})$), low ($I(q_{LF},u_{LF})$) or mixed ($I(q_{MF},u_{MF})$) fidelity models correspondingly map parameters and state from the high, low or mixed fidelity function spaces to QoI values. 
We assume that the parameter and state spaces of the high-, low-, and mixed-fidelity models are such that the same QoI functional can be applied to each to obtain the corresponding QoI values. \red{Don't we also have to assume that the same regularization $R$ and observation $C$ can be applied? does that limit compatibility more than QoI?}
We consider the tradeoff between the error in the QoI and the fidelity (and corresponding computational expense) of the model we use, seeking computable estimates for the error incurred in approximating the high-fidelity model with lower fidelity models. Such an estimate allows us to adaptively choose regions where the high-fidelity model is used, starting from an initial low- or mixed- fidelity model.

@@

%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of models, we take the QoI calculated from inferring the parameters with the highest-fidelity model to be the `exact' or `true' solution; we now derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{proposition}
\label{thm:error_estimate}
Consider the high- and low-fidelity inverse problems described by the constrained optimization problems in \Cref{eq:invOptHF,eq:invOptLF}. Let the forms \red{$a_{HF/LF}:U_{HF/LF} \times U_{HF/LF} \to \Reals$ (doesn't $a$ need three inputs for the output to be a scalar? also what happened to parameter space?)} be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U_{HF/LF}\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q_{HF/LF}\to\Reals$ be differentiable with respect to the parameter $q$, and the Quantity of Interest (QoI) functional $I:Q_{HF/LF}\times U_{HF/LF}\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

>>>>

A given physical system can be described to varying degrees of fidelity using different models. One has a choice in how to define the stucture of the inferred parameters $q$, as well as how to define the state equation and variables. For example, the parameter $q$ can be modeled as a single scalar value or as an infinite-dimensional distributed field variable. The state equation can be nonlinear and multiscale, requiring an expensive computation to resolve (e.g.\ the Navier-Stokes equations) or a linear, single physics model (e.g.\ Laplace's equation). All these different modeling choices lead to different versions of the optimization problem described by \Cref{eq:invOpt} that would need to be solved to infer the unknown parameters.

We assume that there exists an `ideal' model for solving the inverse problem, which we will call the high-fidelity model; given adequate computational resources, we would use this high-fidelity to formulate and solve our inverse problem. We use the subscript $_{HF}$ to denote the high-fidelity inverse problem,
%
\begin{subequations}
\label{eq:invOptHF}
\begin{align}
q_{HF},u_{HF} &=&\arg\min\limits_{q,u} & \quad J(q,u)=\frac{1}{2}\|d-C(u)\|_2^2 + R(q), \label{eq:invOpt_objHF} \\
&&\textrm{s.t. }& \quad a_{HF}(u,q)(\phi)=\ell_{HF}(q)(\phi),\quad\forall\phi\in U_{HF} \label{eq:invOpt_consHF}
\end{align}
\end{subequations}
%
where $q_{HF} \in Q_{HF}$ and $u_{HF}\in U_{HF}$ are the inferred parameters and corresponding state. In practice, one has limited computational resources, so one may need to instead infer the parameters using a less computationally demanding model, which we call the low-fidelity model. We use the subscript $_{LF}$ to denote the low-fidelity inverse problem,
%
\begin{subequations}
\label{eq:invOptLF}
\begin{align}
q_{LF},u_{LF} &=& \arg\min\limits_{q,u} & \quad J(q,u)=\frac{1}{2}\|d-C(u)\|_2^2 + R(q), \label{eq:invOpt_objLF} \\
&& \textrm{s.t. }& \quad a_{LF}(u,q)(\phi)=\ell_{LF}(q)(\phi),\quad\forall\phi\in U_{LF}, \label{eq:invOpt_consLF}
\end{align}
\end{subequations}
%
where $q_{LF} \in Q_{LF}$ and $u_{LF}\in U_{LF}$ are the inferred parameters and corresponding state. If there is compatibility between the low-fidelity and high-fidelity solution spaces for the parameter and state variables (the solutions of one can be interpreted in the context of the other; see \Cref{sec:adapt_alg} for more details), one may combine the two models to form a mixed-fidelity model and corresponding inverse problem. The low-fidelity model can be used in part of the computational domain, and the high fidelity in the remainder; we denote the mixed-fidelity model and its variables with the subscript $_{MF}$. How one chooses to combine the low- and high-fidelity models can be informed by what aspect of the inferred parameters is of interest.

In a {\em goal-oriented inverse problem}, the ultimate purpose of inferring the unknown parameters is to calculate some quantity of interest (QoI). We denote a scalar QoI by a functional that maps the parameters and state to our QoI, $I:Q \times U \to \R$. The QoIs evaluated with the high- ($I(q_{HF},u_{HF})$), low- ($I(q_{LF},u_{LF})$), or mixed- ($I(q_{MF},u_{MF})$) fidelity models correspondingly map parameters and state from the high-, low- or mixed- fidelity function spaces to QoI values. We consider the tradeoff between the error in the QoI and the fidelity (and corresponding computational expense) of the model we use, seeking computable estimates for the error incurred in approximating the high-fidelity model with lower-fidelity models. Such an estimate allows us to adaptively choose regions where the high-fidelity model is used, starting from an initial low- or mixed- fidelity model.

@@

%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of models, we take the QoI calculated from inferring the parameters with the highest-fidelity model to be the `exact' or `true' solution; we now derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{proposition}
\label{thm:error_estimate}
Consider the high- and low-fidelity inverse problems described by the constrained optimization problems in \Cref{eq:invOptHF,eq:invOptLF}. Let the forms $a_{HF/LF}:U_{HF/LF} \times Q_{HF/LF} \times U_{HF/LF} \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U_{HF/LF}\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q_{HF/LF}\to\Reals$ be differentiable with respect to the parameter $q$, and the Quantity of Interest (QoI) functional $I:Q_{HF/LF}\times U_{HF/LF}\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.
